Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-06-20 paper/main.tex
rewrite experiment description

\begin{theorem}{Nyström++ trace estimator for parameter-dependent matrices}{nystrom-pp}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$. If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2 + \log(2\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}} \geq 2$, then for $\delta \in (0, 1/2)$ and $\varepsilon > 0$ with probability at least $1 - \delta$
    \begin{equation}
        \int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t
        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
    \end{equation}
\end{theorem}

@@

\begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
    Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$. Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m \in \mathbb{N}$, $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2 + \log(2\delta^{-1}))$, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
    \begin{equation}
        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t < 2 (1 + \varepsilon) \underline{E}_{\sigma, m} + \varepsilon
        \label{equ:chebyshev-nystrom-error}
        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
        %\label{equ:2-chebyshev-interpolation-error}
    \end{equation}
    for all $\varepsilon > 0$ and $\delta \in (0, 1/2)$.
\end{theorem}

@@

We can express the numerical rank of a matrix with respect to the nuclear norm in terms of its singular values $\sigma_1 \geq \dots \geq \sigma_n \geq 0$
\begin{equation}
    r_{\varepsilon} = \min \{1 \leq r \leq n: \sum_{j=r+1}^n \sigma_{j} \leq \varepsilon \},
    \label{equ:numerical-rank}
\end{equation}
$\varepsilon$ is usually taken to be the double machine precision, i.e. $10^{-16}$. When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ as
\begin{equation}
    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
    \label{equ:gaussian-kernel-eigenvalues}
\end{equation}
where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$. Consequently, by using the closed-form expression of the eigenvalues \refequ{equ:gaussian-kernel-eigenvalues}, we may upper bound the numerical rank of a Gaussian smoothing kernel as
\begin{equation}
    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon, \sigma} \right\},
    \label{equ:gaussian-kernel-numerical-rank}
\end{equation}
with $d_{\varepsilon, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi} \sigma \varepsilon)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants. The expression \refequ{equ:gaussian-kernel-numerical-rank} has a nice visual interpretation: The numerical rank of a matrix is at most equal to the number of eigenvalues which are closer to the parameter $t$ than $d_{\varepsilon, \sigma}$. This is illustrated in \reffig{fig:numerical-rank}.
\begin{figure}[ht]
    \centering
    \input{figures/numerical-rank.tex}
    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ can be
        approximately computed by counting the number of eigenvalues
        $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than
        a constant $d_{\varepsilon, \sigma}$ away from $t$.}
    \label{fig:numerical-rank}
\end{figure}

@@

For our first example, we consider the matrix which arises from the second order
finite difference discretization of the Laplace operator $\Delta$ in a potential
field $V$,
\begin{equation}
    \mathcal{A} u(\vct{x}) = - \Delta u(\vct{x}) + V(\vct{x}) u(\vct{x}),
    \label{equ:5-experiments-electronic-hamiltonian}
\end{equation}
for a uniform mesh of size $h=0.6$. The potential $V$ results from a
lattice whose primitive cell is of side-length $L=6$ and in whose center a
potential
\begin{equation}
    \alpha e^{-\sfrac{\lVert \vct{x} \rVert _2^2}{ 2 \beta^2 }}
    \label{equ:5-experiments-gaussian-cell}
\end{equation}
with $\alpha = -4$, $\beta = 2$ is located. The computational domain is chosen
to span $n_c \in \mathbb{N}$ primitive cells in every spatial dimension, hence, yielding
discretization matrices which are growing in size with $n_c$. In our experiments
we consider the three-dimensional case.%, but for visualization purposes, we illustrate the potential in \reffig{fig:gaussian-well} in two dimensions.

%\begin{figure}[ht]
%    \begin{subfigure}[b]{0.32\columnwidth}
%        \input{plots/gaussian-well-1.pgf}
%        \caption{$n_c=1$}
%        \label{fig:gaussian-well-1}
%    \end{subfigure}
%    \begin{subfigure}[b]{0.32\columnwidth}
%        \input{plots/gaussian-well-2.pgf}
%        \caption{$n_c=2$}
%        \label{fig:gaussian-well-2}
%    \end{subfigure}
%    \begin{subfigure}[b]{0.32\columnwidth}
%        \input{plots/gaussian-well-5.pgf}
%        \caption{$n_c=5$}
%        \label{fig:gaussian-well-5}
%    \end{subfigure}
%    \caption{Two dimensional periodic potential $V$ for different sizes $n_c$ of the computational domain.}
%    \label{fig:gaussian-well}
%\end{figure}

@@

\todo{[Comparison plots with other estimators]}

>>>>

\begin{theorem}{Nyström++ trace estimator for parameter-dependent matrices}{nystrom-pp}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$. If $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1} \log(2\delta^{-1})^2 + \log(2\delta^{-1}))$ with \emph{even} $n_{\mtx{\Omega}}$, then for $\delta \in (0, 1/2)$ and $\varepsilon > 0$ with probability at least $1 - \delta$
    \begin{equation}
        \int_{a}^{b} | \Trace(\mtx{B}(t)) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) |~\mathrm{d}t
        \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
    \end{equation}
\end{theorem}

@@

\begin{theorem}{Chebyshev-Nyström++ method}{chebyshev-nystrom}
    Let $\phi_{\sigma}$ be the smooth spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$. Then the Chebyshev-Nyström++ estimate $\widetilde{\phi}_{\sigma}^{(m)}$ with \emph{even} degree $m \in \mathbb{N}$, $n_{\mtx{\Omega}} = n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-1}\log(2\delta^{-1})^2 + \log(2\delta^{-1}))$, $n_{\mtx{\Omega}}$ \emph{even}, and smoothing parameter $\sigma > 0$ satisfies with probability at least $1 - \delta$
    \begin{equation}
        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \widetilde{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t < 2 (1 + \varepsilon) \underline{E}_{\sigma, m} + \varepsilon
        \label{equ:chebyshev-nystrom-error}
        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
        %\label{equ:2-chebyshev-interpolation-error}
    \end{equation}
    for all $\varepsilon > 0$ and $\delta \in (0, 1/2)$.
\end{theorem}

@@

We can express the numerical rank of a matrix with respect to the nuclear norm in terms of its singular values $\sigma_1 \geq \dots \geq \sigma_n \geq 0$
\begin{equation}
    r_{\varepsilon} = \min \{1 \leq r \leq n: \sum_{j=r+1}^n \sigma_{j} \leq \varepsilon \},
    \label{equ:numerical-rank}
\end{equation}
$\varepsilon$ is usually taken to be the double machine precision, i.e. $10^{-16}$. When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ as
\begin{equation}
    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
    \label{equ:gaussian-kernel-eigenvalues}
\end{equation}
where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$. Consequently, by using the closed-form expression of the eigenvalues \refequ{equ:gaussian-kernel-eigenvalues}, we may upper bound the numerical rank of a Gaussian smoothing kernel as
\begin{equation}
    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A})) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon, \sigma} \right\},
    \label{equ:gaussian-kernel-numerical-rank}
\end{equation}
with $d_{\varepsilon, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi} \sigma \varepsilon)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants. The expression \refequ{equ:gaussian-kernel-numerical-rank} has a nice visual interpretation: The numerical rank of a matrix is at most equal to the number of eigenvalues which are closer to the parameter $t$ than $d_{\varepsilon, \sigma}$. This is illustrated in \reffig{fig:numerical-rank}.
\begin{figure}[ht]
    \centering
    \input{figures/numerical-rank.tex}
    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ can be
        approximately computed by counting the number of eigenvalues
        $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than
        a constant $d_{\varepsilon, \sigma}$ away from $t$.}
    \label{fig:numerical-rank}
\end{figure}

@@

For our first example, we consider the example from \cite{lin-2017-randomized-estimation}, i.e. matrices which arise from the second order finite difference discretization of the Hamiltonian
\begin{equation}
    \mathcal{H} u(\vct{x}) = - \Delta u(\vct{x}) + V(\vct{x}) u(\vct{x}),
    \label{equ:5-experiments-electronic-hamiltonian}
\end{equation}
in a cubic domain for a uniform mesh of size $h=0.6$. The potential $V$ arises from a primitive cell of length $L=6$ which is repeated $n_c \in \mathbb{N}$ times in every spatial dimension, yielding discretization matrices which are growing in size with $n_c$. A Gaussian well
\begin{equation}
    \alpha e^{-\sfrac{\lVert \vct{x} \rVert _2^2}{ 2 \beta^2 }}
    \label{equ:5-experiments-gaussian-cell}
\end{equation}
with $\alpha = -4$, $\beta = 2$ is located in the center of each of these cells. The Hamiltonian models the interaction of a single electron in a crystal. In our experiments we consider the three-dimensional case, but for visualization purposes, we illustrate the potential in \reffig{fig:gaussian-well} in two dimensions.

\begin{figure}[ht]
    \begin{subfigure}[b]{0.32\columnwidth}
        \input{plots/gaussian-well-1.pgf}
        \caption{$n_c=1$}
        \label{fig:gaussian-well-1}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\columnwidth}
        \input{plots/gaussian-well-2.pgf}
        \caption{$n_c=2$}
        \label{fig:gaussian-well-2}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\columnwidth}
        \input{plots/gaussian-well-5.pgf}
        \caption{$n_c=5$}
        \label{fig:gaussian-well-5}
    \end{subfigure}
    \caption{Periodic Gaussian well potential $V$ for different sizes $n_c$ of the computational domain.}
    \label{fig:gaussian-well}
\end{figure}

@@

\todo{[Comparison plots with other estimators?]}
